Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-12 icfp.tex
RM edits end Sec. 2

Finally, we consider default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative. If the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee. We also discuss the special case of a case
expression scrutinizing a variable $x$:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though the program is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
%\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
This interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

>>>>

As for \emph{default} case alternatives, also known as \emph{wildcards}
(written $\_$): matching a wildcard doesn't provide new linearity information,
unlike matching against a pattern with or without linear components, however,
the scrutinee will still be evaluated to WHNF.
%
% so linearity is seen as before but without
% fully consuming the scrutinee linear resources (in non-linear patterns) nor
% binding new linear resources (in linear patterns).
%
Briefly, when matching a wildcard, if the scrutinee is already in WHNF, we can
either use the resources from the scrutinee directly, or the case binder in
that alternative. If the scrutinee is not in WHNF, we \emph{must} use the case
binder, as it's the only way to linearly consume the result of evaluating the
scrutinee.

Finally, we discuss case expressions which scrutinize a
single variable $x$, e.g.:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though this function is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
$x$ returns the result of evaluating $x$ that still had to be consumed exactly
once.
%\end{itemize}
However, in practice, this program's linearity can only be determined by
further branching down on the non-strict evaluation strategy. If linear
function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desireable
practice, as linear functions are known to use their argument exactly once)
% and thus call-by-name reduction will definitely not duplicate computation
while the above function is considered linear, then such an application might
duplicate linear resources during evaluation. For instance:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
On the other hand, if linear function applications are reduced
\emph{call-by-need}, the above example would instead be semantically linear
because the result of freeing $x$ would be bound to some $y$ first.
%
The type system we present in the next section, which evaluates linear
applications call-by-name, does not accept the above function scrutinizing a
variable as linear. 
This interaction between evaluation and linearity comes up in the type
preservation proof of our program, and is again explored with the reverse
binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.
